### 2025-02-12 paper/preprint/paper.tex
small clarification

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically for different classifiers in
Section~\ref{sec-experiments}. Under certain conditions, this results
holds theoretically\footnote{For the proof, see the supplementary
  appendix.}:

@@

Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
\(\mathbf{x} \in \mathbb{R}^D\). If we assume multivariate Gaussian
class densities with common diagonal covariance matrix
\(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), then protecting an
immutable feature from the contrastive divergence penalty
Equation~\ref{eq-div} will result in lower classifier sensitivity to
that feature, provided that at least one other feature is mutable.

@@

By Equation~\ref{eq-loglike-decomp} we have the
\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\)
and \(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\),
where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
\(d\) for class \(k\).

@@

Since \(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) we
are more likely to have
\(x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}} < 0\) than vice
versa at initialization. Similarly, we are more likely to have
\(x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}} > 0\) by
\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\).

This implies that if we do not protect feature \(d_{\text{immtbl}}\),
the contrastive divergence penalty will push down on
\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
effect
\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\).
In words, not protecting the immutable feature would have the
undesirable effect of making the classifier more sensitive to this
feature, in that it would be more likely to predict class \(y^-\) as
opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).

By the same rationale, the contrastive divergence penalty can generally
be expected to push up on \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
\(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\). In
words, this has the effect of making the classifier more sensitive to
the mutable feature, in that it would be more likely to predict class
\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, colframe=quarto-callout-warning-color-frame, opacityback=0, coltitle=black, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, bottomrule=.15mm, colback=white, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, rightrule=.15mm, titlerule=0mm, leftrule=.75mm, opacitybacktitle=0.6, left=2mm]

@@

\subsection{Initial Grid Search}\label{initial-grid-search}

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically for different classifiers in
Section~\ref{sec-experiments}. Under certain conditions, this results
holds theoretically{[}For the proof, see the supplementary appendix.{]}:

@@

Let
\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
denote a linear classifier with softmax activation \(\mathcal{S}\)
(i.e.~\emph{multinomial logistic regression}) where
\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
we assume multivariate Gaussian class densities with common diagonal
covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
then protecting an immutable feature from the contrastive divergence
penalty (Equation~\ref{eq-div}) will result in lower classifier
sensitivity to that feature relative to the remaining features, provided
that at least one of those is mutable.

@@

By the properties of our classifier (\emph{multinomial logistic
regression}), we have:

\begin{equation}\phantomsection\label{eq-multi}{
\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
}\end{equation}

where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
\(d\) for class \(k\).

Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
can identify that
\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
the assumptions we made above. Hence, we have that
\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
and
\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)

@@

Since \((\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})\) we
are more likely to have
\((x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0\) than vice
versa at initialization. Similarly, we are more likely to have
\((x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0\) since
\((\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})\).

This implies that if we do not protect feature \(d_{\text{immtbl}}\),
the contrastive divergence penalty will push down on
\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
effect
\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
In words, not protecting the immutable feature would have the
undesirable effect of making the classifier more sensitive to this
feature, in that it would be more likely to predict class \(y^-\) as
opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).

By the same rationale, the contrastive divergence penalty can generally
be expected to push up on \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
In words, this has the effect of making the classifier more sensitive to
the mutable feature, in that it would be more likely to predict class
\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).

@@

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]

@Cynthia, @Arie, I'm including some preliminary results here but will
rerun experiments in the coming days. You'll notice some odd outlier
results (huge values) for the initial grid search
(Section~\ref{sec-app-initial}). My belief is that this is once again
due to \emph{ECCo} overshooting for some hyperparameter choices, that we
have discussed before. A simple solution to this story is to actually
impose domain constraints. Let's see what the final results show us (I
also still plan to make slight changes to the implementation), but in
any case I think it might even be worth to report results for the
initial grid search in the final appendix (they do include good results
for CT for certain hyperparameter ranges and highlight limitations
inherited from energy-based modelling).

\subsection{Qualitative Findings for Image
Data}\label{qualitative-findings-for-image-data}

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]

@Cynthia, @Arie, Figure~\ref{fig-mnist} shows much more plausible
(faithful) counterfactuals for a model with CT than the model with
conventional training (Figure~\ref{fig-mnist-vanilla}). In fact, this is
not even using \emph{ECCo+} and still showing better results than the
best results we achieved in our AAAI paper for JEM ensembles.

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp.png}}

\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
counterfactual training. The underlying generator, \emph{ECCo}, aims to
generate counterfactuals that are faithful to the model
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp_vanilla.png}}

\caption{\label{fig-mnist-vanilla}Counterfactual images for \emph{MLP}
with conventional training. The underlying generator, \emph{ECCo}, aims
to generate counterfactuals that are faithful to the model
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}

\subsection{Initial Grid Search}\label{sec-app-initial}

### 2025-02-12 paper/sections/appendix.qmd
small clarification

## Initial Grid Search

>>>>

::: {.callout-warning}

\@Cynthia, \@Arie, I'm including some preliminary results here but will rerun experiments in the coming days. You'll notice some odd outlier results (huge values) for the initial grid search (@sec-app-initial). My belief is that this is once again due to *ECCo* overshooting for some hyperparameter choices, that we have discussed before. A simple solution to this story is to actually impose domain constraints. Let's see what the final results show us (I also still plan to make slight changes to the implementation), but in any case I think it might even be worth to report results for the initial grid search in the final appendix (they do include good results for CT for certain hyperparameter ranges and highlight limitations inherited from energy-based modelling).

## Qualitative Findings for Image Data

::: {.callout-note}

\@Cynthia, \@Arie, @fig-mnist shows much more plausible (faithful) counterfactuals for a model with CT than the model with conventional training (@fig-mnist-vanilla). In fact, this is not even using *ECCo+* and still showing better results than the best results we achieved in our AAAI paper for JEM ensembles.

![Counterfactual images for *MLP* with counterfactual training. The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}

![Counterfactual images for *MLP* with conventional training. The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp_vanilla.png){#fig-mnist-vanilla}

## Initial Grid Search {#sec-app-initial}
